The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-29 manuscript/manuscript.Rnw
use the eutherian only core placenta

\begin{figure}[tbh]
<<core.placenta.transcriptome,fig.width=5,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species) & housekeeping==FALSE,]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]

print(ggplot(cptl,aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=9),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(legend.position=c(0.7,0.15),
            axis.text.y=element_text(face="italic"),
            legend.margin=unit(0,"mm"),
            legend.justification="left",
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
  studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. Supplemental
  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different FPKM and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species)
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Min. Exp. in`:=
                                fix.expression.names(`Min. Exp. in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

@@

      + ylab("Number of Genes Meeting FPKM Cutoff in N species")

@@

      + geom_segment(mapping=aes(y=V1,

@@

                         x=300,

@@

      + geom_text(mapping=aes(x=300,
                      y=V1,

@@

                  hjust=0

@@

                                                 `Min. Exp. in` == '14 of 14',

@@

                  hjust=1

@@

      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species"),

@@

                                  label.position="right"

@@

      + theme(legend.position=c(0.98,0.98),

@@

    & `Min. Exp. in` == '14 of 14',max(num_genes)]}

@@

  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
  least 10, %

@@

    & `Min. Exp. in` == '13 of 14',max(num_genes)]}

@@

   with an FPKM of at least 10, and 

@@

    & `Min. Exp. in` == '10 of 14',max(num_genes)]}

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was not
a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

>>>>

\begin{figure}[tbh]
<<core.placenta.transcriptome,fig.width=9/2.5,fig.height=8,out.width="90mm",out.height="0.8\\textheight",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species) & housekeeping==FALSE,]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.41),
            axis.text.y=element_text(face="italic",size=8),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 Eutherians
  studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. Supplemental
  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different FPKM and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=9/2.5,fig.height=9/2.5,out.width="90mm",out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression",
             "all_but_metatherian_expression"
             ),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_but_metatherian_expression$",
              paste0("All ",num.species-1," Eutherians"),
              x
              )
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," Species"),
              x
              )
    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
                   paste0(num.species," of ",num.species," Species")
                   )
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," Species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," Species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Min. Exp. in`:=
                                fix.expression.names(`Min. Exp. in`)]
setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

@@

      + ylab("Number of Genes Meeting FPKM Cutoff")

@@

      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),

@@

                         x=sapply(V1,
                                  function(x){
                                      cpt.distribution.shape[num_genes > x,
                                                             max(`Expression (FPKM)`)*4]
                                  }),

@@

      + geom_text(mapping=aes(x=sapply(V1,
                                       function(x){
                                           cpt.distribution.shape[num_genes > x,
                                                                  max(`Expression (FPKM)`)*4.1]
                                       }),
                      y=V1-ifelse(V1<80,20,0),

@@

                  hjust=0,
                  size=8*(5/14)

@@

                                                 `Min. Exp. in` == 'All 13 Eutherians',

@@

                  hjust=1,
                  size=8*(5/14)

@@

      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),

@@

                                  label.position="right",
                                  label.theme=element_text(size=7,angle=0)

@@

      + theme(text=element_text(size=9),
              legend.position=c(0.99,0.99),

@@

    & `Min. Exp. in` == 'All 13 Eutherians',max(num_genes)]}

@@

  genes would be included if the expression of a set of 1:1 orthologs of all 13 Eutherians must be at
  least 10 FPKM, %

@@

    & `Min. Exp. in` == '13 of 14 Species',max(num_genes)]}

@@

   expression of at least 10 FPKM, and 

@@

    & `Min. Exp. in` == '10 of 14 Species',max(num_genes)]}

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was not
a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.
